\newpage

\begin{center}

{\large \bf ACKNOWLEDGMENTS }

\end{center}

\vspace{1.0cm}

%\setlength{\baselineskip}{0.9 \baselineskip}

It is with great pleasure that I take the time to consider, and to
thank, the many people who have contributed to my life over the past 7
years and to whom I, and this dissertation, are indebted:

First, family.  I thank my parents Stanley and Diana, my sisters Sarah
and Stephanie, my girlfriend Giao, and my brother-in-law Aaron.  Your
enduring love and support, through the difficulties of life and of
graduate school, mean everything to me.

To my thesis committee, without whom I could not have even approached
completing this thesis --\\
%
Frits: you have given me the freedom to find my own winding, branching
road through graduate school.  Always up for discussing whatever crazy
ideas I have had, you have also provided timely and accurate advice,
and been as supportive during rough times as I could possibly have
hoped for.  Thank you.\\
%
Caleb, Zolt\'an, and Kristen: I have enjoyed our collaborations very
much; you have each taught me a great deal about how to be a research
astronomer.
%
To all of you, and to director of graduate studies Jacqueline van
Gorkom: thank you for providing guidance, encouragement, inspiration,
and the occasional nudging.

I am endlessly grateful to Ed Spiegel -- your wisdom and humor in
exchange for my help with your Powerbooks is a trade I'll take any day
of the week, and twice on Sundays.  Thank you, too, Joe Patterson and
David Helfand, without whom I wouldn't have ended up at Columbia in
the first place.  Finally, thank you Norm Baker.  I am honored to have
been a member of the last class you taught.

I thank the graduate students, as colleagues and as friends, with whom
have spent my time here.  In particular, I appreciate the many
scientific conversations and the friendship I have shared with Maurice
Leutenegger, Mark Dijkstra, Michel Zamojski, Tony Mroczkowski, Ben
Sugerman, Aeree Chung, Suvi Gezari, Cameron Hummels, Neil Zimmerman,
Nestor Mirabal, Emily Rauscher, Roban Kramer, Jake Noel-Storr, Jen
Donovan, Alan Gersch.

Thank you, too, non-Columbia friends throughout the years: Matt
Butler, Zimri Yaseen, Mark Macleod, Jeremy Sosenko, Fred Blumberg,
Chris Ingraham, Molly Shearer, Sam Talbot, Joshua Friedman, Jill
Schiff, Eve Armstrong, Karen Schrier.

Millie Garcia: thank you so much for your tremendous, nonjudgmental
help throughout the years with myriad administrative affairs.  You and
Ayoune have made it possible for me to focus almost solely on the job
of being a graduate student.

Other people who have contributed useful ideas, criticism, and advice
to this document include, most significantly, Scott Gaudi, and
additionally Michael Allison, Tony Del Genio, Sara Seager, Kevin Roy
Briggs, George Rybicki, James Cho, David Graff, and several anonymous
referees.

Funding for my graduate school years was provided in part by Columbia
University, by the Columbia Astrobiology Center, and by NASA through
grants NAG5-13354, SAG03-4158A, and Astrobiology: and Planetary
Protection Research grant \#NNG05GO79G.  This research has made use of
publicly available data from the XMM--{\it Newton} Science Archive and
the {\it Swift} observatory data archive.

To everyone else not named above, thank you, too.

%\setlength{\baselineskip}{1.11111 \baselineskip}
